Polynomial Time and Dependent Types

### 2023-10-25 paper/paper.tex
[ paper ] typos and concepts

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2020}
\acmYear{2020}
\acmDOI{10.1145/1122445.1122456}

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

@@

@@

We use an instantiation of QTT with the natural number semiring, with
the usual semiring structure of addition and multiplication. In a mild
generalisation of the original presentation of QTT, we also allow
sub-usaging via the \emph{reverse} ordering on the naturals. That is,
if a variable is marked as usage $n$ and $m \geq n$, then we can also
regard it as usage $m$. This makes the system more like affine linear
logic, since $m \leq 0$\footnote{Reverse ordering!} for all $m$,
matching the system in \autoref{sec:linear-intro}. We do \emph{not}
have an unrestricted usage $\omega$, since this would allow the
possibility of unrestricted duplication, and hence violate our
polytime soundness properties.

@@

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment, where we do not need to care about polytime
realisability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is deciable in polytime. We make three notes
about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
carried out in the $\sigma = 0$ fragment, where we have the full power
of Type Theory to aid us; (ii) this definition is intrinsic, in the
sense that, whichever of the polytime systems is chosen, proving that
a decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (iii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \citet{dallago11realisability}'s realisability models
suitable for modelling dependent types. For a set $A$, the carrier of
$L(A)$ is the set of ternary relations
$X \subseteq A \times M \times \ValSet$ and we define the ordering
$X \leq Y$ to hold iff there exists realising expression
$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $E$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $E$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

The \emph{Cons-free} system uses the $\MaxPoly$ resource monoid, with
the distinguished sub-monoid being those elements that are $0$ in the
size component. We enumerate the features of the \emph{Cons-free}
system and justify their realisability with the $\MaxPoly$ resource
monoid:
\begin{enumerate}
\item Duplication of natural numbers by $\dupNat(M)$ is realisable by
  the expression $(0,0)$, which creates a pair by copying the input
  variable twice. By the cost semantics in
  \autoref{sec:machine-model}, this takes $1$ step of computation (we
  assume that it is actually implemented via some pointer copy).  The
  resource accounting for this realiser works because the size
  component required for the output is the maximum of the size
  components of the two elements, and since $n \sqcup n = n$, we have
  enough resources to fulfil this.
\item Construction of natural numbers is not realisable. In a putative
  $\conSucc$ rule, we would need to get an additional unit of size
  resource from nowhere.
\item Iteration is realised by constructing a realising expression in
  the expression language from the given expressions for the zero and
  successor cases that uses the in-built recursion of the
  language. The proof that resources are correctly accounted for is
  carried out by induction on the natural being iterated over. For
  $n$, we require potential
  $\mathit{scale}(n,\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, where
  $\gamma_{\mathit{succ}}$ and $\gamma_{\mathit{zero}}$ are the
  potentials required by the successor and zero cases respectively. By
  Property (2) of iteration resource monoids, above, we know that
  $\mathit{raise}(\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$ always
  dominates this requirement when paired with the potential
  $\mathit{size}(n)$ from the input. Therefore, this latter
  expression, plus some administrative set up costs, is the required
  potential for the whole iterator.
\end{enumerate}
Together, we have a soundness result for the \emph{Cons-free} system,
that ensures that every term in the $\sigma = 1$ fragment is
realisable by a \emph{correct} program that terminates in polynomial
time for all inputs:
\begin{theorem}[Soundness for the Cons-free System]
  \label{thm:cons-free-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n)$ and $v$ is
  a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
\end{theorem}

@@

Soundness for the LFPL system is similar to the \emph{Cons-free}
system, except for a $+1$ to the input to the polynomial, to account
for the fact that we cost one size unit for the $\conZero$
constructor.
\begin{theorem}[Soundness for the LFPL-style System]
  \label{thm:lfpl-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n+1)$ and $v$
  is a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket X \rrbracket(n)$.
\end{theorem}

@@

Our construction already includes soundness of a system with intrinsic
but explicit resource tracking where $\Diamond$s are used to pay for
every step of computation but never returned, via the natural number
resource monoid defined in
\autoref{sec:specific-monoids}. Investigation of such a system may
yield a system that tracks intrinsic costs precisely.

@@

The realisability type $\mathbf{R}(A)$ described in
\autoref{sec:qtt-reflection} allows us to internalise the
realisability of certain functions into the logical ($\sigma = 0$)
fragment of the calculus. However, it is not possible to derive any
logical consequences from this other than turning it back into a
function. This limitation becomes acute when trying to prove results
from standard Computational Complexity theory. Even though we can
characterise the class NP, as we did in \autoref{sec:np-class}, and it
is a ``matter of programming'' to show that 3-SAT is in NP, we cannot
prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
because the proofs relies on obtaining the \emph{source code} of the
program solving an NP problem and then encoding that program in
3-SAT. To do this in our setting, we would need to internalise the
soundness property (\thmref{thm:lfpl-soundness}) as an axiom, stating
that for a realisable polytime function there (merely) exists a
realising expression $E$ that completes in polynomial time, and then
writing polytime enodings into 3-SAT. We hope that the addition of
such an axiom to our system would lead to an expressive machine-free
\emph{Synthetic Computational Complexity Theory}, analogous to the
Church-Turing axiom for Synthetic Computability Theory \cite{Bauer06}.

@@

\section{Typing Rules for QTT}

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{10.1145/1122445.1122456}

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10003801</concept_id>
       <concept_desc>Theory of computation~Linear logic</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003777.10003778</concept_id>
       <concept_desc>Theory of computation~Complexity classes</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10003777.10003787</concept_id>
       <concept_desc>Theory of computation~Complexity theory and logic</concept_desc>
       <concept_significance>300</concept_significance>
       </concept>
   <concept>
       <concept_id>10003752.10010124.10010131.10010137</concept_id>
       <concept_desc>Theory of computation~Categorical semantics</concept_desc>
       <concept_significance>100</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Linear logic}
\ccsdesc[500]{Theory of computation~Type theory}
\ccsdesc[500]{Theory of computation~Complexity classes}
\ccsdesc[300]{Theory of computation~Complexity theory and logic}
\ccsdesc[100]{Theory of computation~Categorical semantics}

@@

In this section, we give an overview of the rules of QTT.

The full rules are presented in \autoref{sec:rules}.

The full rules are presented in the online appendix at

@@

We use an instantiation of QTT with the natural number semiring, with
the usual semiring structure of addition and multiplication. In a mild
generalisation of the original presentation of QTT, we also allow
sub-usaging via the \emph{reverse} ordering on the naturals. That is,
if a variable is marked as usage $n$ and $m \geq n$, then we can also
regard it as usage $m$. This makes the system more like affine linear
logic, since $m \sqsubseteq 0$\footnote{Reverse ordering!} for all $m$,
matching the system in \autoref{sec:linear-intro}. We do \emph{not}
have an unrestricted usage $\omega$, since this would allow the
possibility of unrestricted duplication, and hence violate our
polytime soundness properties.

@@

$\Sigma$-types are a little more involved, and demonstrate the
flexibility in QTT in allowing additional power in the $\sigma = 0$
fragment where we do not need to care about polytime
realisability. Formation and introduction are given by the rules:
\begin{mathpar}
  \inferrule*
  {0\Gamma \vdash S~\istype \\ 0\Gamma, x \stackrel0: S \vdash T~\istype}
  {0\Gamma \vdash (x \stackrel\pi: S) \otimes T~\istype}

@@

We can use the reflection type former defined in
\autoref{sec:qtt-reflection} to define a predicate on decision
problems that establishes whether or not they are polytime decision
problems. Specifically, we can state that there is a polytime
realisable boolean-value predicate that reports true exactly when the
given element of $a$ is in the predicate:
\begin{displaymath}
  \mathrm{PTIME}(A,P) =
    (f \stackrel1: \mathbf{R}(A \to \BoolTy)) \otimes (a \stackrel1: A) \to (\mathbf{R}^{-1}(f)\, a = \cTrue) \Leftrightarrow P\,a
\end{displaymath}
Thus, $\mathrm{PTIME}(A,P)$ is a logical proposition stating that the
decision problem $(A,P)$ is decidable in polytime. We make three notes
about this definition: (i) \emph{proofs} of $\mathrm{PTIME}(A,P)$, are
carried out in the $\sigma = 0$ fragment, where we have the full power
of Type Theory to aid us; (ii) this definition is intrinsic, in the
sense that, whichever of the polytime systems is chosen, proving that
a decision problem is solvable in polytime is a matter of programming,
without having to reason directly about machine models and step
counting; and (iii) we have defined problems to have arbitrary types
$A$ as domains, rather than bitstrings, and so the notion of size
attached to an input is intrinsic to the type $A$ chosen.

@@

We now define an indexed linear poset $L$ of realisers over $\Set$
that ties together our ``mathematical'' model of types in $\Set$ with
our machine model and resource monoid. This construction is a
reformulation of \citet{dallago11realisability}'s realisability models
suitable for modelling dependent types. For a set $A$, the carrier of
$L(A)$ is the set of ternary relations
$X \subseteq A \times M \times \ValSet$ and we define the ordering
$X \leq Y$ to hold iff there exists a realising expression
$E \in \ExpSet$ and potential $\gamma \in M_0$ such that for all
$a \in A$, $\alpha \in M$ and $v \in \ValSet$ with
$(a,\alpha,v) \in X$, we have that there exists a result
$v' \in \ValSet$, step count $k \in \mathbb{N}$ and result potential
$\beta \in M$ with:
\begin{enumerate}
\item $E, v \Downarrow_k v'$ (evaluation successfully completes in $k$ steps);
\item $(a, \beta, v') \in Y$ (the result is well-resourced and
  satisfies $Y$); and
\item $k \leq M(\alpha \rplus \gamma, \beta)$ (the step count is within the
  difference between the initial potential and the result potential).
\end{enumerate}
Note that the definition of realisablity is uniform in the element $a$
-- the realising expression $E$ and the potential $\gamma$ must work
for all $a$ -- thus the implementation and complexity measure of the
transition being modelled cannot depend on what the input is. Put in
implementation terms, the input $a$ is not present at
runtime. Moreover note that the potential $\gamma$ attached to the
expression $E$ must come from the sub-monoid $M_0$, indicating that is
intended to be data-free, while the potential $\alpha$ for the input
is from the full monoid $M$, so it can contain data and functions.

@@

The \emph{Cons-free} system uses the $\MaxPoly$ resource monoid, with
the distinguished sub-monoid being those elements that are $0$ in the
size component. We enumerate the features of the \emph{Cons-free}
system and justify their realisability with the $\MaxPoly$ resource
monoid:
\begin{enumerate}
\item Duplication of natural numbers by $\dupNat(M)$ is realisable by
  the expression $(0,0)$, which creates a pair by copying the input
  variable twice. By the cost semantics in
  \autoref{sec:machine-model}, this takes $1$ step of computation (we
  assume that it is actually implemented via some pointer copy).  The
  resource accounting for this realiser works because the size
  component required for the output is the maximum of the size
  components of the two elements, and since $n \sqcup n = n$, we have
  enough resources to fulfil this.
\item Construction of natural numbers is not realisable. In a putative
  $\conSucc$ rule, we would need to get an additional unit of size
  resource from nowhere.
\item Iteration is realised by constructing a realising expression in
  the expression language from the given expressions for the zero and
  successor cases that uses the in-built recursion of the
  language. The proof that resources are correctly accounted for is
  carried out by induction on the natural being iterated over. For
  $n$, we require potential
  $\mathit{scale}(n,\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$, where
  $\gamma_{\mathit{succ}}$ and $\gamma_{\mathit{zero}}$ are the
  potentials required by the successor and zero cases respectively. By
  Property (2) of iteration resource monoids, above, we know that
  $\mathit{raise}(\mathit{acct}(4) \rplus \gamma_{\mathit{succ}})
  \rplus (\mathit{acct}(2) \rplus \gamma_{\mathit{zero}})$ always
  dominates this requirement when paired with the potential
  $\mathit{size}(n)$ from the input. Therefore, this latter
  expression, plus some administrative set up costs, is the required
  potential for the whole iterator.
\end{enumerate}
Together, we have a soundness result for the \emph{Cons-free} system,
that ensures that every term in the $\sigma = 1$ fragment is
realisable by a \emph{correct} program that terminates in polynomial
time for all inputs:
\begin{theorem}[Soundness for the Cons-free System]
  \label{thm:cons-free-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n)$ and $v$ is
  a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

@@

Soundness for the LFPL system is similar to the \emph{Cons-free}
system, except for a $+1$ to the input to the polynomial, to account
for the fact that we cost one size unit for the $\conZero$
constructor.
\begin{theorem}[Soundness for the LFPL-style System]
  \label{thm:lfpl-soundness}
  If we have a term $n : \tyNat \vdash M \stackrel1: T(n)$ then there
  exists a realising expression $E$ and polynomial $p$ such that for
  all $n \in \mathbb{N}$, there exists $v \in \ValSet$ and
  $k \in \mathbb{N}$ such that
  $E, [\mathrm{natValue}(n)] \Downarrow_k v$, $k \leq p(n+1)$ and $v$
  is a realising value for
  $\llbracket M \rrbracket(n) \in \llbracket T \rrbracket(n)$.
\end{theorem}

@@

Our construction already includes soundness of a system with intrinsic
but explicit resource tracking where $\Diamond$s are used to pay for
every step of computation but never returned, via the natural number
resource monoid defined in
\autoref{sec:specific-monoids}. Investigation of such a system may
yield a system that tracks the intrinsic cost of programs precisely
and explicitly.

@@

The realisability type $\mathbf{R}(A)$ described in
\autoref{sec:qtt-reflection} allows us to internalise the
realisability of certain functions into the logical ($\sigma = 0$)
fragment of the calculus. However, it is not possible to derive any
logical consequences from this other than turning it back into a
function. This limitation becomes acute when trying to prove results
from standard Computational Complexity theory. Even though we can
characterise the class NP, as we did in \autoref{sec:np-class}, and it
is a ``matter of programming'' to show that 3-SAT is in NP, we cannot
prove the Cook-Levin theorem that 3-SAT is NP-complete. This is
because the proofs relies on obtaining the \emph{source code} of the
program solving an NP problem and then encoding that program in
3-SAT. To do this in our setting, we would need to internalise the
soundness property (\thmref{thm:lfpl-soundness}) as an axiom, stating
that for a realisable polytime function there (merely) exists a
realising expression $E$ that completes in polynomial time, and then
writing polytime encodings into 3-SAT. We hope that the addition of
such an axiom to our system would lead to an expressive machine-free
\emph{Synthetic Computational Complexity Theory}, analogous to the
Church-Turing axiom for Synthetic Computability Theory \cite{Bauer06}.

@@

\section{Typing Rules for QTT/Cons-free and QTT/LFPL}
\label{sec:rules}
